LLM Output Compliance with Handcrafted Linguistic Features --- An Experiment

### 2024-12-10 article/main.tex
reword section 3.1 and make other minor corrections

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
These values are usually easy to compute with methods that are idempotent.
This makes HLFs very attractive because of their dual potential.
They can be used to instruct the generation of text (``write no more than 10
sentences''), but they can also be used to analyse text.
This allows us to verify if a text has the \textit{expected} value for an HLF.\@

Taking Noam Chomsky's ideas that syntax and meaning are different
things~\cite{chomsky2002syntactic}, we intuit that writing style is defined in
terms of both.
We shy away from discussing meaning, but syntax and similarly orderly constructs
seem to be within our grasp.
Traditionally, linguistic style revolved around morphological and syntactic
arrangements~\cite{lugea2023stylistics}.
More recent descriptions find semantics and pragmatics to be inextricably linked
to author style~\cite{verma2019lexical}, but argue that a great deal can still
be inferred from lexicon, syntax and a `surface' level which contains basic HLFs
such as the number of words in a sentence.
Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

@@

Linguistic style is the sum of identifiable language choices manifest in a text
made from the language system by the text producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
Opinions seem to converge that style depends on context and author choice
vis-a-vis a communication goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent also from a
formalist perspective~\cite{chomsky2002syntactic}.
Our selection of HLFs is informed by the various aspects of linguistic style.

@@

Our experiment analyses LLM output when asked to generate new text based on
given input text and instructions derived from HLFs.
We compute statistics of HLFs on a corpus to derive the text generation
instructions.
To run the experiment we must select LLMs to evaluate and find datasets from
which we can derive corpora.
Then we must also select the HLFs for our experiment.
Before going over our selection for the experiment's parameters, let us go
through the experimental process.

@@

Assume we have a target LLM, a corpus containing texts, some examples extracted
from the corpus, a selection of HLFs and input text both from within and from
outside the corpus.
This description of the experiment process assumes those parameters have stable
values.
Changing the parameters requires restarting the process.

The first activity involves establishing baselines for each experiment scenario.
In the first scenario the target LLM is instructed to generate text retaining
the meaning of the input text from outside the corpus without any further
instruction related to style or the linguistic features of the generated text.
This step is repeated 10 times, computing each HLF in our selection for the
generated text.
The computed values represent the baseline HLF values for the first scenario.

The second scenario consists in generating text based on input text, but with
the aid of some examples from the corpus.
The examples are appended to the instruction prompt used in the first scenario.
The number of examples is chosen so that we do not overrun the context window of
the LLM.\@
Some context windows are smaller than others, setting the number of examples at
three.
Adding examples to the LLM's instruction prompt also makes the provenance of the
input text more important.
Therefore we experiment with two variants of the second scenario: one with input
text extraneous to the corpus and one with input text from the corpus.
The text from outside the corpus is the same text we used in the first scenario.
The text from within the corpus is chosen randomly once at the start of the
experiment.
We repeat the same process from the first scenario to obtain baseline HLF values
for each variant of the second scenario.

From here on we refer to `baseline HLF values' as `baseline'.
We now have three baselines: one for the first scenario and one for each variant
of the second scenario.

The next step requires computing statistics on the corpus for each HLF:\@
\textbf{min}~(the minimum value), \textbf{max}~(the maximum value) and
\textbf{avg}~(the mean).
These statistics allow writing an HLF prompt derived from each of our baseline
prompts.

The HLF prompts instruct the LLM to generate text with HLFs that comply with the
previously computed statistics.
The individual (per-HLF) instructions are all part of the same HLF prompt.
The prompts are constructed using the Jinja2~\cite{jinja2} templates which are
located in the \texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/src/llm_tst_consistency/templates}{templates}}
directory of the source code used to run the experiment.
The instruction prompts corresponding to our two scenarios are
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The rest of the templates in the directory are used to construct the individual
HLF instructions.

In a similar manner to how we generated the baselines, we run 10 trials for each
scenario and variant.
We compare each baseline with the results obtained by using the corresponding
HLF prompt.
From here on we refer to the results obtained by using prompts with HLF
instructions as `HLF results'.

The first scenario indicates whether LLMs acknowledge HLF instructions.
If they do, the HLF results should be significantly different from their
baseline.

Both variants of the second scenario quantify how similar the HLFs of the texts
generated by the LLM are to the corpus average \textit{because of} the HLF
instructions.
Working under the assumption that an LLM can generate text that is similar to an
example from the corpus, the presence of HLF instructions in the prompt would
remain among the few possible explanations were we to observe deviations from
the baseline.

@@

We only use state of the art large language models for the experiment.
The Chatbot Arena~\cite{chiang2024chatbot} leaderboard is essential in our
selection.
To cover more ground we find top performers from different vendors interesting.
The final selection is available in Table~\ref{table-llm}.

@@

Our main focus is to understand how susceptible LLMs are to prompts containing
instructions to conform to HLFs.
LFTK~\cite{lftk-2023} is a framework built on top of spaCy~\cite{spacy}
that provides implementations for computing many HLFs.
We use these two libraries to perform our experiment.
LFTK categorizes HLFs by domain and family and make available information on the
suitability of each HLF for given tasks such as readability assessment or fake
news detection.
Besides the domain and family, we also choose features based on their influence
over text style as recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

The instructions for each HLF are specifically crafted using natural language.
The resulting instructions can be catalogued on the concrete-abstract spectrum.
For example, `write 100 words' is a more concrete instruction than `write as if
you were in junior highschool'.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category.
Table~\ref{table-hlf} showcases our selection of features.
The LFTK ID is used throughout tables and figures to refer to a specific HLF.\@

@@

The experiment design rests upon measuring HLFs on a corpus of texts and
computing statistics.
LLMs also make use of examples extracted from a corpus to guide text generation.
Our approach idempotently derives a corpus of texts from a well known dataset.
Our selection of datasets is listed in Table~\ref{table-ds}.

@@

CNN/DailyMail contains stories and news with diverse styles.
The style follows the publisher's guidelines, but, within those guidelines, it
varies for each author.
The articles are more formal, longer and more complex in structure than Yelp
reviews which makes this dataset a good alternative.
The corpus that is based on this dataset contains the first 1000 CNN articles
from the test split of the dataset.

\subsection{External Input Texts}\label{input-text}

The experiment involves using input text from both outside the corpus and from
within it.
This section describes how we use text external to the corpus, having already
described our usage of text from within the corpus.

We use a different external input text for each corpus.
These texts are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's source code.

The text external to the CNN article corpus is an extract from the classic
\textit{The Life and Opinions of Tristram Shandy, Gentleman}, by Laurence
Sterne~\cite{sterne2003life}.
The text style is very dissimilar compared to news articles.
Its linguistic features are also very unlike the average values computed for the
CNN article corpus.
A text that is similarly divorced from the style and linguistic features of most
Yelp reviews is Oscar Wilde's \textit{Sonnet to Liberty}~\cite{wilde1909poems}.
Both texts are in the public domain and were obtained from Project
Gutenberg~\cite{gutenberg}.

@@

Handcrafted linguistic features are integer or floating point values that we
can compute on text.
Plot figures show the HLF on the vertical axis and the trial number on the horizontal
axis.
We plot the baseline using a continuous grey line and the HLF result using a
continuous grenat line.
The minimum and maximum HLF on the corpus are plotted using dotted black lines.
A continuous black line designates the average HLF on the corpus.
We refer to the corpus average as the \textit{target} because the LLM is
instructed to generate text that primarily conforms to this value.

@@

We can determine whether the baseline results differ from the HLF results using
the Kolmogorov-Smirnov test~\cite{kolmogorov1933,smirnov1939}, as implemented in
the Scipy library~\cite{2020SciPy-NMeth}.
We consider that the results are different for a $p\_value \le 0.05$.

To determine which of two results is closer to the target, we compute the area
between each result and the target.
Another way in which we measure the closeness of the result to the target is by
computing the Euclidian norm of the vectorial difference between the result and
target.

Note that each HLF can have a different range of values than other HLFs.
When we compute the area and the Euclidian norm using Numpy~\cite{harris2020array},
we use the raw values of each HLF, without preprocessing them.
Therefore the results for one HLF are not comparable with the results for
another HLF.\@
Since we don't quantify how different the baseline is from the result and we
compute the relative closeness to a common target, comparing values for the same
HLF but different LLMs is also ill-advised.

Let us denote with $A_b$ and $A_h$ the areas between baseline and target and HLF
result and target, respectively.
Let us similarly denote with $N_b$ and $N_h$ the Euclidian norms.
We define $Diff_A = A_b - A_h$ and $Diff_N = N_b - N_h$.

The result tables show $Diff_A$ and $Diff_N$ for each LLM in the context of a
prompt and a dataset.
All result tables for each dataset and prompt variant are available in the
appendix.

Next, we present our perspective on the most relevant results of the experiment.

@@

\begin{table*}[ht!]
    \centering
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{lllllllllll}
            model                 & t\_word                      & t\_sent                    & n\_uverb                    & n\_uadj                    & simp\_ttr                 & a\_verb\_pw               & corr\_adj\_var            & corr\_verb\_var            & fkgl                    & a\_kup\_pw                \\
            \toprule
            claude3 $Diff_N$     & \textbf{\underline{623.16}}  & \textbf{\underline{24.74}} & \textbf{\underline{51.18}}  & \textbf{\underline{26.68}} & 0.06                      & -0.0                      & \textbf{\underline{0.46}} & \textbf{\underline{1.4}}   & 1.44                    & \textbf{\textit{-1.66}}   \\ \midrule
            claude3 $Diff_A$     & \textbf{\underline{1774.0}}  & \textbf{\underline{74.0}}  & \textbf{\underline{158.84}} & \textbf{\underline{81.2}}  & 0.2                       & -0.01                     & \textbf{\underline{1.88}} & \textbf{\underline{4.49}}  & 5.01                    & \textbf{\textit{-4.09}}   \\ \midrule
            gemini $Diff_N$      & \textbf{\underline{985.86}}  & \textbf{\underline{34.52}} & \textbf{\underline{100.36}} & \textbf{\underline{53.39}} & \textbf{\underline{0.52}} & 0.05                      & \textbf{\underline{2.62}} & \textbf{\underline{4.68}}  & -4.17                   & \textbf{\textit{-2.43}}   \\ \midrule
            gemini $Diff_A$      & \textbf{\underline{2990.5}}  & \textbf{\underline{106.5}} & \textbf{\underline{320.34}} & \textbf{\underline{159.1}} & \textbf{\underline{1.5}}  & 0.09                      & \textbf{\underline{7.24}} & \textbf{\underline{13.15}} & -12.65                  & \textbf{\textit{-7.13}}   \\ \midrule
            gpt $Diff_N$         & \textbf{\underline{125.87}}  & 1.46                       & 10.95                       & 10.77                      & 0.02                      & -0.01                     & 0.71                      & 0.81                       & \textbf{\textit{-1.72}} & 0.43                      \\ \midrule
            gpt $Diff_A$         & \textbf{\underline{327.0}}   & 3.0                        & 29.0                        & 27.0                       & 0.08                      & -0.01                     & 1.82                      & 2.24                       & \textbf{\textit{-5.72}} & 0.44                      \\ \midrule
            command\_r $Diff_N$  & \textbf{\underline{1024.78}} & \textbf{\underline{39.67}} & \textbf{\underline{81.58}}  & \textbf{\underline{41.46}} & \textbf{\underline{0.46}} & \textbf{\underline{0.03}} & \textbf{\underline{2.67}} & \textbf{\underline{3.78}}  & 1.4                     & \textbf{\underline{0.14}} \\ \midrule
            command\_r $Diff_A$  & \textbf{\underline{3132.5}}  & \textbf{\underline{120.5}} & \textbf{\underline{246.84}} & \textbf{\underline{132.2}} & \textbf{\underline{1.36}} & \textbf{\underline{0.06}} & \textbf{\underline{7.88}} & \textbf{\underline{11.21}} & 3.9                     & \textbf{\underline{1.03}} \\ \midrule
            llama3\_70b $Diff_N$ & \textbf{\underline{459.45}}  & \textbf{\underline{17.78}} & \textbf{\underline{36.5}}   & \textbf{\underline{19.75}} & \textbf{\underline{0.12}} & \textbf{\underline{0.03}} & \textbf{\underline{1.43}} & \textbf{\underline{2.18}}  & 1.5                     & \textbf{\textit{-0.06}}   \\ \midrule
            llama3\_70b $Diff_A$ & \textbf{\underline{1363.5}}  & \textbf{\underline{54.0}}  & \textbf{\underline{106.5}}  & \textbf{\underline{60.5}}  & \textbf{\underline{0.36}} & \textbf{\underline{0.09}} & \textbf{\underline{4.33}} & \textbf{\underline{6.38}}  & 7.46                    & \textbf{\textit{-0.11}}   \\ \bottomrule
        \end{tabular}%
    }\caption{CNN Results}\label{table-prompt-1-cnn-dailymail}
\end{table*}

Figure~\ref{fig-command-r-t-word} shows a positive result, where the result
obtained with the aid of HLF instructions is closer to the target.
Figure~\ref{fig-gpt-fkgl} shows a negative result, where the baseline is closer
to the target.
Figure~\ref{fig-llama-simp-ttr} shows an inconclusive result.

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-command_r-cnn_dailymail/prompt_1-command_r-cnn_dailymail_t_word.png}
        \caption{Command-R+\\Total Words}\label{fig-command-r-t-word}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-gpt-cnn_dailymail/prompt_1-gpt-cnn_dailymail_fkgl.png}
        \caption{GPT\\Flesch-Kincaid Grade Level}\label{fig-gpt-fkgl}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-claude3-cnn_dailymail/prompt_1-claude3-cnn_dailymail_a_verb_pw.png}
        \caption[center]{Claude3\\Average Verbs per Word}\label{fig-llama-simp-ttr}
    \end{minipage}
\end{figure*}

@@

We attempt quantifying the impact of using HLF instructions on the output of the
LLM by eliminating as many factors as possible that would lead to a different
text being generated when compared to a baseline.
In the second scenario of our experiment we use examples to allow the LLM to
generate better baselines, thus hopefully narrowing the gap between baseline
and HLF-enhanced result.

There are two variants of this scenario depending on the provenance of the input
text.
The input text may be from outside the corpus or a text from within the corpus.
Table~\ref{table-prompt-2-cnn-dailymail} contains results for the first variant:
input from outside the corpus.
We notice the HLF results tend to be closer to the target than the baseline
results.
This picture is similar to the one we saw in the first scenario.

>>>>

\textbf{Handcrafted linguistic features (HLFs)} are single numerical values
produced by a uniquely identifiable method on any natural
language~\cite{lftk-2023}.
Examples of HLFs range from simple constructs such as counts or averages of
words, sentences or specific parts of speech~(adjectives, verbs, a.s.o) to more
complex statistics based on heuristics.
All HLFs share the characteristics of computational ease and idempotence.
This makes HLFs very attractive because of their dual potential.
On the one hand, an LLM might be instructed using a prompt containing the value
of an HLF.\@
An example prompt instruction which asks the LLM to conform to a maximum
\textit{sentence count} of 10 would read ``write no more than 10 sentences''.
On the other hand, HLFs can be computed on the text output by an LLM, too.
This dual nature of HLFs allows inspecting how closely does LLM generated text
conform to the goal outlined in the input prompt HLF instruction.\@

If syntax differs from meaning~\cite{chomsky2002syntactic} the intuition is that
writing style is defined in terms of both.
The distinction is important in order to set boundaries.
HLFs are largely focused on syntax and similarly orderly constructs.
Therefore, we cover the influence of meaning on author style minimally and
superficially at best.

This limitation should not deter readers because, traditionally, linguistic
style is centered around morphological and syntactic
arrangements~\cite{lugea2023stylistics}.
Although recent descriptions find semantics and pragmatics to be inextricably
linked to author style~\cite{verma2019lexical}, they too argue that a great deal
can still be inferred from lexicon, syntax and a `surface' level which contains
basic HLFs such as the number of words in a sentence.
Some have observed that many characteristics of written text can be computed as
HLFs~\cite{hovy1987generating,lugea2023stylistics}.
Combining multiple HLFs that are revelatory of linguistic style could provide us
with an easily interpretable and verifiable stylistic profile.

@@

Linguistic style is the sum of identifiable language choices manifest in a text
made from the language system by the text producer~\cite{lugea2023stylistics}.
Another way to think about it is that style is the form used for delivering
meaning~\cite{tst_sigkdd_review_2022}.
Opinions seem to converge that style depends on context and author choice
vis-a-vis a communication goal~\cite{mcdonald1985computational,hovy1987generating}.
The author's choice is available at all linguistic levels, including the
morphological, lexical, syntactic, semantic and pragmatic
levels~\cite{dimarco1994model,lugea2023stylistics}.
The influence of syntactic structure over text style is apparent from a formal
perspective~\cite{chomsky2002syntactic}.
Our selection of HLFs is informed by the various aspects of linguistic style.

@@

Our experiment aims to find out how well LLMs can follow prompts which have
instructions derived from HLFs.
Two scenarios are investigated.

The aim in the first scenario is to use an LLM to reword an input text so that
its style resembles the text style of texts from a predetermined corpus of
similarly styled texts.
In this scenario the LLM rewords the input by following instructions which only
contain HLFs.
The input text used for this scenario is necessarily from outside the corpus.

The second scenario is designed to show how much bearing do instructions
containing HLFs have on the way LLMs generate text.
The task of the LLM remains the same as in the first scenario.
However, examples from the chosen corpus are added to the LLM prompt which means
the LLM rewords text aided by the examples.
The second scenario has two variants because of this particularity.
In the first variant, similarly to the first scenario, the LLM rewords input
from outside the corpus.
In the second variant of the second scenario the LLM uses input text from within
the corpus.

@@

The experiment starts with choosing experiment parameter values.
The following parameters have stable values throughout the experiment process: a
selection of HLFs, a corpus containing texts, a target LLM, examples from the
chosen corpus, one input text from the chosen corpus and one input text from
outside the chosen corpus.

Once the parameters have been determined, the HLF statistics are computed on the
text corpus.
Remembering that HLFs are just floating point numbers, the following statistics
are computed for each HLF:~\textbf{min}~(the minimum value), \textbf{max}~(the
maximum value) and \textbf{avg}~(the mean).

The subsequent step involves formulating prompt instructions that incorporate
HLFs.
For each HLF, an instruction is constructed in natural language, adhering to the
previously calculated minimum, maximum, and mean values. For instance, an
instruction corresponding to the total word count HLF might be expressed as
follows:

\texttt{- Total Words: ensure the text contains 14 words (min=10, max=50).}

The HLF instructions are used to build system prompts for the LLM.\@
The prompts are assembled using Jinja2~\cite{jinja2} templates located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/src/llm_tst_consistency/templates}{templates}}
directory of the experiment's source code.

The prompt templates correspond to each of the two experiment scenarios:
`\texttt{prompt\_1}' and `\texttt{prompt\_2}', respectively.
The static instruction in `\texttt{prompt\_1}' to ``\textit{Write text that
conveys the meaning of the first user prompt}'' ensures that input text meaning
is preserved.
In addition to the primary directive from `\texttt{prompt\_1}',
`\texttt{prompt\_2}' contains instructions concerning the interpretation of the
provided examples, as well as logic to incorporate them.
Both prompt templates contain logic to distinguish between the situation when
HLF instructions were given versus not.

With the preliminary steps completed, baseline HLFs for the target LLM are
computed.
For the first scenario, the LLM is instructed to reword the input text 10 times
by using `\texttt{prompt\_1}' without HLF instructions.
The HLFs in our selection are computed for each generated text, resulting in
a 10-dimensional vector for each HLF.\@
These vectors represent the baseline HLFs for the LLM in the first scenario.

The second scenario baseline values are computed similarly using the
`\texttt{prompt\_2}' template and the examples provided as one of the experiment
parameters.
Remembering that the second scenario has two variants, we obtain two sets of
baseline vectors: a set for input from the corpus and one for input outside the
corpus.
The input from outside the corpus is the same as the input for the first
scenario.

From here on we refer to the HLFs computed using the above process as
\textit{baselines}.
Note that we have 3 sets of baselines: 1 for the first scenario and 2 for the
second scenario.

For each type of baseline, we run 10 text generations using the corresponding
prompt augmented with the HLF instructions computed earlier in the process.
Doing this results in 3 sets of 10-dimensional vectors, each corresponding to
the baselines.
We refer to these vectors as \textit{HLF results}.

The last step is to compare baselines with their corresponding HLF results for
each HLF in our selection, for each scenario and variant.

The comparison result for the first scenario indicates whether LLMs acknowledge
HLF instructions at all.
If they do, the HLF results should be significantly different from their
baseline.

Both variants of the second scenario quantify how similar the HLFs of the texts
generated by the LLM are to the corpus average \textit{because of} the HLF
instructions.
This hypothesis emerges from the assumption that LLMs are able to mimic examples
when they are provided in the system prompt.
In such a case, the presence of HLF instructions would remain among the few
possible explanations for deviations from the baseline.

@@

The LLMs selected for the experiment must represent the state of the art.
The Chatbot Arena~\cite{chiang2024chatbot} leaderboard is essential in our
selection based on this criterion.
To cover more ground we find top performers from different vendors interesting.
The final selection is available in Table~\ref{table-llm}.

@@

Our main focus is to understand how susceptible LLMs are to prompts containing
instructions to conform to HLFs.
In this experiment we use LFTK~\cite{lftk-2023}, a framework built on top of
spaCy~\cite{spacy} that provides implementations for computing many HLFs.
LFTK categorizes HLFs by domain and family.
We select HLFs from the catalog which is implemented by LFTK based on their
domain and family, with the interest to make a diverse selection with respect to
these attributes.
Besides the domain and family, we also choose features based on their influence
over text style as recognized in literature~\cite{verma2019lexical,lugea2023stylistics}.

The instructions for each HLF are specifically crafted using natural language.\@
The resulting instructions can be catalogued on the concrete-abstract spectrum.
For example, `write 100 words' is a more concrete instruction than `write as if
you were in junior highschool'.
We divide HLFs in five categories ranging from very concrete to highly abstract
and choose 2 features from each category.

Table~\ref{table-hlf} showcases our HLF selection.
The LFTK ID is used throughout tables and figures to refer to a specific HLF.\@

@@

A corpus of texts is required to perform the experiment.
We have conducted the experiment using corpora derived idempotently from the
datasets listed in Table~\ref{table-ds}.

The main criteria for choosing these datasets were the difference in style
between the texts contained in one corpus versus the other and the similarity in
style between the texts from the same corpus.

@@

CNN/DailyMail contains stories and news with diverse styles.
The style follows the publisher's guidelines, but, within those guidelines, it
varies slightly for each author.
The articles are more formal, longer and more complex in structure than Yelp
reviews which makes this dataset a good alternative.
The corpus that is based on this dataset contains the first 1000 CNN articles
from the test split of the dataset.

\subsection{Input Text Choice}\label{input-text}

The experiment involves using input text from both outside the corpus and from
within it.

An input text from within each corpus is chosen so that it exhibits the closest
HLFs to the averages computed on the corpus.
The choice is made when the text corpora are selected and is stable throughout
all experimental evaluations.

External input text is chosen in relation to each corpus.
The chosen texts are located in the
\texttt{\textlinkanon{https://github.com/koosie0507/llm-tst-consistency/tree/main/data}{data}}
folder of the experiment's source code.

The text external to the CNN article corpus is an extract from the classic
\textit{The Life and Opinions of Tristram Shandy, Gentleman}, by Laurence
Sterne~\cite{sterne2003life}.
The text style is not similar to the style of news articles at all.
Our selected HLFs are also very different from the average values computed for
the CNN article corpus.
A text that is similarly divorced from the style and HLFs of most Yelp reviews
is Oscar Wilde's \textit{Sonnet to Liberty}~\cite{wilde1909poems}.
Both texts are in the public domain and were obtained from Project
Gutenberg~\cite{gutenberg}.

@@

HLFs are real numbers that we can derive from text idempotently.
Plot figures show the HLF on the vertical axis and the trial number on the horizontal
axis.
We plot the baseline using a continuous grey line and the HLF result using a
continuous grenat line.
The minimum and maximum HLF on the corpus are plotted using dotted black lines.
A continuous black line designates the average HLF on the corpus.
We refer to the corpus average as the \textit{target} because the LLM is
instructed to generate text that primarily conforms to this value.

@@

Because HLFs are just real numbers highlighting characteristics of text, they
can be used for setting goals to aim for (`write a 100 word paragraph') and for
describing text attributes (`this paragraph has 100 words').
In our experiments we set goals for the chosen LLMs using instruction prompts
and compute the corresponding HLFs on the generated text in order to measure
LLM efficacy.

As outlined in Subsection~\ref{subsec:experiment-process}, our experiment
compares a baseline with the corresponding HLF results.
Remember that in the context of a scenario variant, both the baseline and the
HLF results for a specific HLF are vectors containing 10 real numbers.

The Kolmogorov-Smirnov test~\cite{kolmogorov1933,smirnov1939} is useful for
determining whether the baseline and HLF results vectors differ significantly.
We use the Scipy~\cite{2020SciPy-NMeth} implementation of the test.
In our experiment, the vectors are significantly different for a
$p\_value \le 0.05$.

For each HLF, the target HLF value is repeated 10 times to construct a vector
which is compatible with the baselines and HLF results.
This allows us to compute the Euclidian norm of the vectorial difference between
the baselines and HLF results on one hand and the target vector, on the other.
A similar computation can be performed based on the area between each of the
vectors and the target.
Because the area measurements are correlated with the Euclidian norm, we present
results using only the Euclidian norm.

Note that each HLF can have a different range of values than other HLFs.
While computing the Euclidian norm using Numpy~\cite{harris2020array}, we use
the raw values of each HLF, without preprocessing.
Therefore the norm computed for one HLF does not imply anything in relation to
another HLF or LLM.\@

Let us denote with $N_b$ and $N_h$ the Euclidian norms for the baseline and the
HLF results, respectively.
We denote the difference between the norms with $Diff_N = N_b - N_h$.

The result tables show $Diff_N$ for each LLM in the context of a prompt and a
text corpus.
While the main body of the paper is focused on the CNN dataset and, an appendix
to the paper documents the results collected by running the experiment for each
dataset and prompt variant.

@@

\begin{table*}[ht!]
    \centering
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{lllllllllll}
            model       & t\_word                      & t\_sent                    & n\_uverb                    & n\_uadj                    & simp\_ttr                 & a\_verb\_pw               & corr\_adj\_var            & corr\_verb\_var            & fkgl                    & a\_kup\_pw                \\
            \toprule
            claude3     & \textbf{\underline{623.16}}  & \textbf{\underline{24.74}} & \textbf{\underline{51.18}}  & \textbf{\underline{26.68}} & 0.06                      & -0.0                      & \textbf{\underline{0.46}} & \textbf{\underline{1.4}}   & 1.44                    & \textbf{\textit{-1.66}}   \\ \midrule
            gemini      & \textbf{\underline{985.86}}  & \textbf{\underline{34.52}} & \textbf{\underline{100.36}} & \textbf{\underline{53.39}} & \textbf{\underline{0.52}} & 0.05                      & \textbf{\underline{2.62}} & \textbf{\underline{4.68}}  & -4.17                   & \textbf{\textit{-2.43}}   \\ \midrule
            gpt         & \textbf{\underline{125.87}}  & 1.46                       & 10.95                       & 10.77                      & 0.02                      & -0.01                     & 0.71                      & 0.81                       & \textbf{\textit{-1.72}} & 0.43                      \\ \midrule
            command\_r  & \textbf{\underline{1024.78}} & \textbf{\underline{39.67}} & \textbf{\underline{81.58}}  & \textbf{\underline{41.46}} & \textbf{\underline{0.46}} & \textbf{\underline{0.03}} & \textbf{\underline{2.67}} & \textbf{\underline{3.78}}  & 1.4                     & \textbf{\underline{0.14}} \\ \midrule
            llama3\_70b & \textbf{\underline{459.45}}  & \textbf{\underline{17.78}} & \textbf{\underline{36.5}}   & \textbf{\underline{19.75}} & \textbf{\underline{0.12}} & \textbf{\underline{0.03}} & \textbf{\underline{1.43}} & \textbf{\underline{2.18}}  & 1.5                     & \textbf{\textit{-0.06}}   \\ \midrule
        \end{tabular}%
    }\caption{$Diff_N$ on the CNN-DailyMail corpus\\Scenario 1}\label{table-prompt-1-cnn-dailymail}
\end{table*}

Figure~\ref{fig-command-r-t-word} shows a positive result, where the result
obtained with the aid of HLF instructions is closer to the target.
Figure~\ref{fig-gpt-fkgl} shows a negative result, where the baseline is closer
to the target.
Figure~\ref{fig-claude3-avpw} shows an inconclusive result.

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-command_r-cnn_dailymail/prompt_1-command_r-cnn_dailymail_t_word.png}
        \caption{Command-R+~\texttt{t\_word}\\CNN-DailyMail,~Scenario 1}\label{fig-command-r-t-word}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-gpt-cnn_dailymail/prompt_1-gpt-cnn_dailymail_fkgl.png}
        \caption{GPT~\texttt{fkgl}\\CNN-DailyMail,~Scenario 1}\label{fig-gpt-fkgl}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_1/prompt_1-claude3-cnn_dailymail/prompt_1-claude3-cnn_dailymail_a_verb_pw.png}
        \caption[center]{Claude3~\texttt{a\_verb\_pw}\\CNN-DailyMail,~Scenario 1}\label{fig-claude3-avpw}
    \end{minipage}
\end{figure*}

@@

The second scenario of our experiment uses examples to allow the LLM to
generate better baselines.
As pointed out at the end of Subsection~\ref{subsec:experiment-process}, we
assume that the difference we measure between baselines and HLF results is due
to the HLF instructions in the system prompt.

The first variant of the second scenario involves using input text from outside
the corpus.
Table~\ref{table-prompt-2-cnn-dailymail} contains the measurements we obtained
in this context.
We notice the HLF results tend to be closer to the target than the baseline
results.
In terms of significance and efficacy of the HLF instructions, the results are
similar to the ones obtained in the first scenario.

\begin{table*}[ht!]
    \centering
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{lllllllllll}
            model       & t\_word                      & t\_sent                    & n\_uverb                    & n\_uadj                    & simp\_ttr                 & a\_verb\_pw               & corr\_adj\_var            & corr\_verb\_var           & fkgl                       & a\_kup\_pw                \\
            \toprule
            claude3     & \textbf{\underline{718.49}}  & \textbf{\underline{24.13}} & \textbf{\underline{44.43}}  & \textbf{\underline{23.43}} & \textbf{\underline{0.4}}  & 0.03                      & 0.94                      & \textbf{\underline{0.77}} & 1.2                        & 0.07                      \\ \midrule
            gemini      & \textbf{\underline{785.6}}   & \textbf{\underline{25.8}}  & \textbf{\underline{60.5}}   & \textbf{\underline{41.6}}  & \textbf{\underline{0.41}} & \textbf{\underline{0.05}} & \textbf{\underline{1.95}} & \textbf{\underline{2.67}} & -4.62                      & \textbf{\textit{-1.94}}   \\ \midrule
            gpt         & 47.8                         & 5.18                       & 6.13                        & -0.13                      & 0.02                      & 0.0                       & 0.05                      & 0.3                       & 3.06                       & -0.27                     \\ \midrule
            command\_r  & \textbf{\underline{263.51}}  & 5.35                       & \textbf{\underline{25.08}}  & \textbf{\underline{17.37}} & \textbf{\textit{-0.09}}   & -0.01                     & \textbf{\underline{1.74}} & \textbf{\underline{1.42}} & \textbf{\underline{10.46}} & \textbf{\underline{1.56}} \\ \midrule
            llama3\_70b & \textbf{\underline{376.0}}   & \textbf{\underline{15.63}} & \textbf{\underline{38.63}}  & 15.74                      & \textbf{\underline{0.24}} & \textbf{\underline{0.06}} & 1.69                      & \textbf{\underline{2.91}} & 0.31                       & -0.3                      \\ \midrule
        \end{tabular}%
    }
    \caption{$Diff_N$ on the CNN-DailyMail corpus\\Scenario 2, Input Outside Corpus}\label{table-prompt-2-cnn-dailymail} % chktex-file 8
\end{table*}

We should note that there are fewer significant differences between baseline and
HLF results.
This seems to validate our assumption that using examples in the system prompt
causes the baselines to be closer to the target HLF value.

\begin{figure*}[ht!]
    \centering
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-gpt-cnn_dailymail/prompt_2-gpt-cnn_dailymail_t_word.png}
        \caption{GPT~\texttt{t\_word}\\CNN-DailyMail, Scenario 2, Input Outside Corpus}\label{fig-p2-gpt-twords}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-command_r-cnn_dailymail/prompt_2-command_r-cnn_dailymail_simp_ttr.png}
        \caption{Command-R+~\texttt{simp\_ttr}\\CNN-DailyMail, Scenario 2, Input Outside Corpus}\label{fig-p2-command-r-simpttr}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \includegraphics[width=\linewidth]{plots/prompt_2/prompt_2-gemini-cnn_dailymail/prompt_2-gemini-cnn_dailymail_a_kup_pw.png}
        \caption[center]{Gemini~\texttt{a\_kup\_pw}\\CNN-DailyMail, Scenario 2, Input Outside Corpus}\label{fig-p2-gemini-a-kup-pw}
    \end{minipage}
\end{figure*}
